Enzo: An Adaptive Mesh Refinement Code for Astrophysics

### 2013-06-24 numerical-radiation-fld.tex
Changes from Cameron Hummels comments (also some other typos I found while working through his comments).

In addition to the ray-tracing approach for radiation transport
described in Section~\ref{sec.num.raytracing}, \enzo\ currently includes
a field-based radiation transport solver for problems posed on uniform
(i.e.~non-AMR, non-SMR) grids, which has been tuned for large-scale simulations
involving many ionizing sources.  Detailed explanations of the model and
solution approach may be found in
\citet{NBHBROW2007}, \citet{ReynoldsHayesPaschosNorman2009},
\citet{NRS2009}, and \citet{2013arXiv1306.0645N}, the salient features of which are reproduced here. 
In addition, comparisons of this solver with other astrophysical
radiation transport solvers may be found in \cite{IlievEtAl2009}.  
\enzo's field-based radiation solver focuses on
a flux-limited diffusion approximation for cosmological radiative
transfer, with couplings to both the gas energy and chemical number
densities.
%\begin{eqnarray}
%  \label{eq:fld_radiation}
%  \partial_t E_r + \frac1a \div \left(E_r \myvec{v}_b\right) &=& 
%    \div\left(D\grad E_r\right) -
%    \frac{\dot{a}}{a} E_r - c\kappa E_r + \eta, \\
%  \label{eq:fld_heating}
%  \partial_t e_c &=& -\frac{2\dot{a}}{a} e_c + G - \Lambda, \\
%  \label{eq:fld_chemistry}
%  \partial_t {\mathrm n}_i + 
%    \frac1a \div\left({\mathrm n}_i \myvec{v}_b\right) &=& 
%    \alpha_{i,j} {\mathrm n}_e {\mathrm n}_j - {\mathrm n}_i
%    \Gamma_i^{ph}, \quad i=1,\ldots,N_s,
%\end{eqnarray}
%where $E_r$ is a grey radiation energy density, $e_c$ is the
%internal energy correction due to photo-heating and chemical cooling,
%and ${\mathrm n}_i$ correspond to the chemical number densities of
%HI, HII, HeI, HeII and HeIII, and ${\mathrm n}_e$ is the electron
%number density.  Here, we define the grey radiation energy density
%through first assuming a fixed frequency spectrum, i.e.
%$E_{\nu}(\nu,\myvec{x},t) = \tilde{E}_r(\myvec{x},t) \chi(\nu)$, and then
%defining the integrated quantity
%\begin{equation}
%\label{eq:grey_radiation_energy}
%   E_r(\myvec{x},t) \equiv \int_{\nu_0}^{\infty}
%   E_{\nu}(\nu,\myvec{x},t)\,\mathrm d\nu \  = \ 
%   \tilde{E}_r(\myvec{x},t) \int_{\nu_0}^{\infty} \chi(\nu)\,\mathrm d\nu.
%\end{equation}
%In addition, $D$ in (\ref{eq:fld_radiation}) is the {\em Larsen
%square-root flux-limiter} (see \cite{Morel2000}), $\kappa$ is the total
%opacity, $\eta$ is the field of radiation sources, $G$ provides the
%radiation induced photo-heating, $\Lambda$ provides the chemical
%cooling, $\alpha_{i,j}$ are the reaction rate coefficients defining
%the interactions between species ${\mathrm n}_i$ and ${\mathrm n}_j$,
%and $\Gamma_i^{ph}$ are the radiation induced photo-ionization rates.

>>>>

In addition to the ray-tracing approach for radiation transport
described in Section~\ref{sec.num.raytracing}, \enzo\ currently includes
a field-based radiation transport solver for problems posed on uniform
(i.e.~non-AMR, non-static mesh refinement) grids, which has been tuned for large-scale simulations
involving many ionizing sources.  Detailed explanations of the model and
solution approach may be found in
\citet{NBHBROW2007}, \citet{ReynoldsHayesPaschosNorman2009},
\citet{NRS2009}, and \citet{2013arXiv1306.0645N}, the salient features of which are reproduced here. 
In addition, comparisons of this solver with other astrophysical
radiation transport solvers may be found in \cite{IlievEtAl2009}.  
\enzo's field-based radiation solver focuses on
a flux-limited diffusion approximation for cosmological radiative
transfer, with couplings to both the gas energy and chemical number
densities.
%\begin{eqnarray}
%  \label{eq:fld_radiation}
%  \partial_t E_r + \frac1a \div \left(E_r \myvec{v}_b\right) &=& 
%    \div\left(D\grad E_r\right) -
%    \frac{\dot{a}}{a} E_r - c\kappa E_r + \eta, \\
%  \label{eq:fld_heating}
%  \partial_t e_c &=& -\frac{2\dot{a}}{a} e_c + G - \Lambda, \\
%  \label{eq:fld_chemistry}
%  \partial_t {\mathrm n}_i + 
%    \frac1a \div\left({\mathrm n}_i \myvec{v}_b\right) &=& 
%    \alpha_{i,j} {\mathrm n}_e {\mathrm n}_j - {\mathrm n}_i
%    \Gamma_i^{ph}, \quad i=1,\ldots,N_s,
%\end{eqnarray}
%where $E_r$ is a grey radiation energy density, $e_c$ is the
%internal energy correction due to photo-heating and chemical cooling,
%and ${\mathrm n}_i$ correspond to the chemical number densities of
%HI, HII, HeI, HeII and HeIII, and ${\mathrm n}_e$ is the electron
%number density.  Here, we define the grey radiation energy density
%through first assuming a fixed frequency spectrum, i.e.
%$E_{\nu}(\nu,\myvec{x},t) = \tilde{E}_r(\myvec{x},t) \chi(\nu)$, and then
%defining the integrated quantity
%\begin{equation}
%\label{eq:grey_radiation_energy}
%   E_r(\myvec{x},t) \equiv \int_{\nu_0}^{\infty}
%   E_{\nu}(\nu,\myvec{x},t)\,\mathrm d\nu \  = \ 
%   \tilde{E}_r(\myvec{x},t) \int_{\nu_0}^{\infty} \chi(\nu)\,\mathrm d\nu.
%\end{equation}
%In addition, $D$ in (\ref{eq:fld_radiation}) is the {\em Larsen
%square-root flux-limiter} (see \cite{Morel2000}), $\kappa$ is the total
%opacity, $\eta$ is the field of radiation sources, $G$ provides the
%radiation induced photo-heating, $\Lambda$ provides the chemical
%cooling, $\alpha_{i,j}$ are the reaction rate coefficients defining
%the interactions between species ${\mathrm n}_i$ and ${\mathrm n}_j$,
%and $\Gamma_i^{ph}$ are the radiation induced photo-ionization rates.

### 2013-06-24 numerical-starformation.tex
Jeff Oishi's changes.
Changes from Cameron Hummels comments (also some other typos I found while working through his comments).

Due to the computationally unfeasible number of stars in a galaxy
($10^{11}$) and the lack of detailed understanding of star formation, a
number of phenomenological star formation models are included in
\enzo.
%The \enzo\ code includes approximately ten models for star formation
%and feedback.  
Broadly speaking, these methods all work in similar
ways: At a specified time interval, all grid cells that are at the
highest local level of refinement (i.e., that have no child cells) are
examined to see if they meet a set of criteria for star formation.
This may simply be a baryon density threshold, but may include more
complex tests, such as an examination of local cooling and dynamical
time scales, molecular hydrogen fraction, metallicity, and converging
gas flows.  If the star formation criteria are met, some mass of gas
is taken away from the cell in question and a ``star particle'' with
the same mass is placed in the center of that cell, with the same
velocity as the removed gas.  This star particle is then
allowed to inject mass, momentum, thermal energy, metals, and possibly
magnetic fields and/or cosmic ray populations into its local
environment.  In general, the particle is treated as an ensemble of
stars, with feedback properties occuring over time according to the
assumed initial mass function of the stellar population.

@@

The Cen \& Ostriker method is a heuristic model of star formation on
galactic scales.  This method, first described by \citet{CO1992},
assumes that stars form in substantially overdense, converging, and
gravitationally unstable gas.  Algorithmically, every cell at the
locally highest level of refinement are examined at every timestep to
see if they meet the critiera for star formation, and star particles
are allowed to form in a cell if the following criteria are met:
\begin{eqnarray}
\rho_b/\bar{\rho}_b & \geq & \eta,  \\[2mm]
\label{cendens}
%\end{equation} 
%\begin{equation}
\div \vec{v}_b & < & 0, \\
\label{cencont}
%\end{equation}
%\begin{equation}
t_{\rm cool} & < & t_{\rm dyn} \equiv \sqrt{3 \pi / 32G \rho_{\rm tot}}, \\
%\end{equation}
%\begin{equation}
m_{b} & > & m_{J} \equiv G^{-3/2} \rho_{b}^{-1/2}c^{3} 
\left[ 1 + \frac{\delta\rho_{d}}{\delta\rho_{b}} \right]^{-3/2}
\end{eqnarray}
where 
%$\rho_b$ is the baryon density in that cell, $\bar{\rho_b}$ is
%the mean baryon density in the simulation, 
$\eta$ is the user-defined
overdensity threshold, 
%$\vec{v}$ is the velocity of the gas in the
%cell, $\rho_{d}$ is the dark matter density, $\rho_{\rm tot} =
%\rho_{b} + \rho_{d}$, 
and $m_{b}$ and $m_{j}$ are the baryonic mass in the
cell and the Jeans mass of the cell, and c is the isothermal sound speed
in the cell.  If all of these criteria are met, the mass of a star
particle is calculated as \(m_{*} = m_{b} \frac{ \Delta t}{ t_{\rm
    dyn} } f_{\rm *eff} \), where $f_{\rm *eff}$ is the star formation
efficiency parameter.

@@

\item A converging gas flow ($\div \vec{v} < 0$), and

>>>>

Due to the computationally unfeasible number of stars in a galaxy
($10^{11}$) and the lack of detailed understanding of star formation, a
number of phenomenological star formation models are included in
\enzo.
%The \enzo\ code includes approximately ten models for star formation
%and feedback.  
Broadly speaking, these methods all work in similar
ways: At a specified time interval, all grid cells that are at the
highest local level of refinement (i.e., that have no child cells) are
examined to see if they meet a set of criteria for star formation.
This may simply be a baryon density threshold, but may include more
complex tests, such as an examination of local cooling and dynamical
time scales, molecular hydrogen fraction, metallicity, and converging
gas flows.  If the star formation criteria are met, some mass of gas
is taken away from the cell in question and a ``star particle'' with
the same mass is placed in the center of that cell with the same
velocity as the removed gas.  This star particle is then
allowed to inject mass, momentum, thermal energy, metals, and possibly
magnetic fields and/or cosmic ray populations into its local
environment.  In general, the particle is treated as an ensemble of
stars, with feedback properties occuring over time according to the
assumed initial mass function of the stellar population.

@@

The Cen \& Ostriker method is a heuristic model of star formation on
galactic scales.  This method, first described by \citet{CO1992},
assumes that stars form in substantially overdense, converging, and
gravitationally unstable gas.  Algorithmically, each cell at the
locally highest level of refinement is examined at each timestep to
see if it meets the critiera for star formation.  Star particles
are allowed to form in a cell if the following criteria are met:
\begin{eqnarray}
\rho_b/\bar{\rho}_b & \geq & \eta,  \\[2mm]
\label{cendens}
%\end{equation} 
%\begin{equation}
\div \myvec{v}_b & < & 0, \\
\label{cencont}
%\end{equation}
%\begin{equation}
t_{\rm cool} & < & t_{\rm dyn} \equiv \sqrt{3 \pi / 32G \rho_{\rm tot}}, \\
%\end{equation}
%\begin{equation}
m_{b} & > & m_{J} \equiv G^{-3/2} \rho_{b}^{-1/2}c^{3} 
\left[ 1 + \frac{\delta\rho_{d}}{\delta\rho_{b}} \right]^{-3/2}
\end{eqnarray}
where 
%$\rho_b$ is the baryon density in that cell, $\bar{\rho_b}$ is
%the mean baryon density in the simulation, 
$\eta$ is the user-defined
overdensity threshold, 
%$\myvec{v}$ is the velocity of the gas in the
%cell, $\rho_{d}$ is the dark matter density, $\rho_{\rm tot} =
%\rho_{b} + \rho_{d}$, 
and $m_{b}$ and $m_{j}$ are the baryonic mass in the
cell and the Jeans mass of the cell, and c is the isothermal sound speed
in the cell.  If all of these criteria are met, the mass of a star
particle is calculated as \(m_{*} = m_{b} \frac{ \Delta t}{ t_{\rm
    dyn} } f_{\rm *eff} \), where $f_{\rm *eff}$ is the star formation
efficiency parameter.

@@

\item A converging gas flow ($\div \myvec{v} < 0$), and

### 2013-06-24 numerical-zeus.tex
Changes from Cameron Hummels comments (also some other typos I found while working through his comments).

As an alternative to the previous Godunov methods, \enzo\ also includes an implementation of the finite difference hydrodynamic algorithm employed in the compressible magnetohydrodynamics code \zeus\ \citep{Stone92a, Stone92b}.  Fluid transport is solved on a Cartesian grid using the upwind, monotonic advection scheme of \citet{1977JCoPh..23..276V} within a multistep (operator split) solution procedure that is fully explicit in time.  This method is formally second-order-accurate in space but first-order-accurate in time.

As discussed in the section describing the Piecewise Parabolic Method (Section~\ref{sec.hydro.ppm}), operator split methods break the solution of the hydrodynamic equations into parts, with each part representing a single term in the equations.  Each part is evaluated successively using the results preceding it.  In this method, in addition to operator splitting the expansion terms (i.e. those terms in eqs.~(\ref{eq:mass}) -- (\ref{eq:total_energy}) that depend on $\dot{a}$), we divide the remaining terms into \emph{source} and \emph{transport} steps.  The terms to be solved in the source step are those on the right-hand side of eqs.~(\ref{eq:mass}) -- (\ref{eq:total_energy}), while the transport terms are on the left-hand side of these equations and are responsible for the advection of mass, momentum and energy across the grid.

>>>>

As an alternative to the previous Godunov methods, \enzo\ also includes an implementation of the finite difference hydrodynamic algorithm employed in the compressible magnetohydrodynamics code \zeus\ \citep{Stone92a, Stone92b}.  Fluid transport is solved on a Cartesian grid using the upwind, monotonic advection scheme of \citet{1977JCoPh..23..276V} within a multistep (operator-split) solution procedure that is fully explicit in time.  This method is formally second-order accurate in space but first-order accurate in time.

As discussed in the section describing the Piecewise Parabolic Method (Section~\ref{sec.hydro.ppm}), operator-split methods break the solution of the hydrodynamic equations into parts, with each part representing a single term in the equations.  Each part is evaluated successively using the results preceding it.  In this method, in addition to operator-splitting the expansion terms (i.e. those terms in eqs.~(\ref{eq:mass}) -- (\ref{eq:total_energy}) that depend on $\dot{a}$), we divide the remaining terms into \emph{source} and \emph{transport} steps.  The terms to be solved in the source step are those on the right-hand side of eqs.~(\ref{eq:mass}) -- (\ref{eq:total_energy}), while the transport terms are on the left-hand side of these equations and are responsible for the advection of mass, momentum and energy across the grid.

### 2013-06-24 overview.tex
Changes from Cameron Hummels comments (also some other typos I found while working through his comments).

The primary feature of the \enzo\ code is its Adaptive Mesh Refinement
(AMR) capability, which allows it to reach extremely large dynamical
ranges with limited computational resources, opening doors to applications otherwise
closed by finite memory and computational time. Unlike moving mesh
methods \citep{1995ApJS..100..269P,1995ApJS...97..231G} or methods
that subdivide individual cells \citep{Adjerid}, Berger \& Collela's
AMR (also referred to as \emph{structured} AMR, or SAMR) utilizes an adaptive
hierarchy of grid patches at varying levels of resolution.  Each
rectangular grid patch (referred to as a ``grid'') covers some region
of space in its \emph{parent grid} requiring higher resolution,
and can itself become the parent grid to an even more highly resolved
\emph{child grid}.

@@

We begin with the direct-Eulerian PPM implementation.  This is an
explicit, higher-order-accurate version of Godunov's method for ideal
gas dynamics with the spatially third-order-accurate piecewise parabolic monotonic
interpolation and a nonlinear Riemann solver for shock capturing.  It
advances the hydrodynamic equations in the following steps:
%\begin{enumerate}
% \item 
(i) Construct monotonic parabolic interpolation of cell average data,
for each fluid quantity;
% \item 
(ii) Compute interface states by averaging the parabola over the
domain of dependence for each interface;
% \item 
(iii) Use interface data to solve the Riemann problem;
 %\item 
(iv) Difference the interface fluxes to update the cell average
quantities.
%\end{enumerate}

The PPM implementation does an excellent job of capturing strong
shocks across a few cells.  Multidimensional schemes are built up by
directional splitting and produce a method that is formally second-order-accurate in space and time and which explicitly conserves mass,
linear momentum, and energy \citep{Hawley84, Norman86}.  A variety of
Riemann solvers have been implemented.

>>>>

The primary feature of the \enzo\ code is its Adaptive Mesh Refinement
(AMR) capability, which allows it to reach extremely large dynamical
ranges with limited computational resources, opening doors to applications otherwise
closed by finite memory and computational time. Unlike moving mesh
methods \citep{1995ApJS..100..269P,1995ApJS...97..231G} or methods
that subdivide individual cells \citep{Adjerid}, Berger \& Collela's
AMR (also referred to as \emph{structured} AMR, or SAMR) utilizes an adaptive
hierarchy of grid patches at varying levels of resolution.  Each
rectangular grid patch (referred to as a ``grid'') covers some region
of space requiring higher resolution than its \emph{parent grid},
and can itself become the parent grid to an even more highly resolved
\emph{child grid}.

@@

We begin with the direct-Eulerian PPM implementation.  This is an
explicit, higher-order accurate version of Godunov's method for ideal
gas dynamics with the spatially third-order accurate piecewise parabolic monotonic
interpolation and a nonlinear Riemann solver for shock capturing.  It
advances the hydrodynamic equations in the following steps:
%\begin{enumerate}
% \item 
(i) Construct monotonic parabolic interpolation of cell average data,
for each fluid quantity;
% \item 
(ii) Compute interface states by averaging the parabola over the
domain of dependence for each interface;
% \item 
(iii) Use interface data to solve the Riemann problem;
 %\item 
(iv) Difference the interface fluxes to update the cell average
quantities.
%\end{enumerate}

The PPM implementation does an excellent job of capturing strong
shocks across a few cells.  Multidimensional schemes are built up by
directional splitting and produce a method that is formally
second-order accurate in space and time and which explicitly conserves mass,
linear momentum, and energy \citep{Hawley84, Norman86}.  A variety of
Riemann solvers have been implemented.

### 2013-06-24 table_of_variables_and_constants.tex
Changes from Cameron Hummels comments (also some other typos I found while working through his comments).

  \startdata
  \multicolumn{2}{l}{\underline{Fluid Quantities}} \\[5pt]
  $E$ & comoving total fluid energy density \\
  $e$ & comoving thermal energy density \\
  $T$ & baryon temperature \\
  $S$ & fluid entropy \\
  $\rho$ & comoving baryonic density \\
  $\rho_{\!X}$ & mass density of component $X$ (gas, dark matter, stars, metals, total, etc.) \\
  $t$ & time \\
  $\vecv$ & peculiar velocity \\
  $\vecB$ & comoving magnetic field strength \\
  $p$ & thermal pressure \\
  $p^*$ & total (thermal + magnetic) isotropic pressure \\
  $\gamma$ & ideal gas ratio of specific heats \\
  $\phi$ & gravitational potential \\
  $\Lambda$ & radiative/chemical cooling rate \\
  $\Gamma$ & radiative heating rate \\
  $Q$ & artificial viscosity stress tensor (only for ZEUS hydro) \\[3pt]

>>>>

  \startdata
  \multicolumn{2}{l}{\underline{Fluid Quantities}} \\[5pt]
  $E$ & comoving total fluid energy density \\
  $e$ & comoving thermal energy density \\
  $T$ & baryon temperature \\
  $S$ & fluid entropy \\
  $\rho$ & comoving baryonic density \\
  $\rho_{\!X}$ & comoving mass density of component $X$ (gas, dark matter, stars, metals, total, etc.) \\
  $t$ & time \\
  $\vecv$ & peculiar velocity \\
  $\vecB$ & comoving magnetic field strength \\
  $p$ & thermal pressure \\
  $p^*$ & total (thermal + magnetic) isotropic pressure \\
  $\gamma$ & ideal gas ratio of specific heats \\
  $\phi$ & gravitational potential \\
  $\Lambda$ & radiative/chemical cooling rate \\
  $\Gamma$ & radiative heating rate \\
  $Q$ & artificial viscosity stress tensor (only for ZEUS hydro) \\[3pt]
